\chapter{More on geometric algebra}

Although vector geometric algebra provides a simple, convenient model for embedding our
understanding of vectors into the framework of geometric algebra, it is by no means the only model.
Models such as \textit{projective geometric algebra} (PGA) and \textit{conformal geometric algebra}
(CGA) use extra dimensions to assist in the modeling of various kinds of objects in space.

\section{Familiar number systems are geometric algebras}

We've previously seen how the 2D vector geometric algebra contains the complex numbers in its even
subalgebra. The complex numbers themselves are a geometric algebra, $\ClR{0,1}$. Here,
we've extended the Clifford algebra symbol to include the number of dimensions that square to a
negative number - that dimension is the imaginary numbers, with the real numbers being part of the
scalar field that always squares to 1.

Similarly, the quaternions also comprise a geometric algebra, $\ClR{0,2}$. It's easy to show that
the two vectors and the bivector of the space map onto the quaternionic elements $i$, $j$, and $k$
(the order of correspondence is irrelevant).

Unfortunately, we can't go higher. The octonions aren't associative, violating a fundamental
requirement that Clifford algebras hold - they are unital associative algebras. The Cayley-Dickson
contstruction allows for the construction of algebras over the field of real numbers with
increasingly loose constraints as the number of dimensions increases.

\section{Projective geometric algebra}

A \textit{projective geometric algebra} (PGA) is a geometric algebra that contains one or more extra
projective dimensions. These dimensions are \textit{projective}, meaning that they are essentially
flattened into the remaining dimensions. When we calculate a length, components of a multivectors
that contain the projective dimensions square to zero.

We will extend our descriptions of the underlying Clifford algebra by adding a third dimension index
representing the components that square to zero. The n-dimensional projective geometric algebra is
modeled by the Clifford algebra $\ClR{n,0,1}$.

Projective geometric algebra provides a framework for treating translations and rotations
equivalently. This makes it an extremely useful model for working with space groups - the symmetries
that are possible in crystals. It also serves as a foundation for 3D computer graphics - the
projective properties naturally describe the process of flattening a 3D object into a 2D
representation.

\section{Conformal geometric algebra}

\section{Properties of geometric algebras}

Every geometric algebra - in other words, every real Clifford algebra - satisfies some common
properties that are worth pointing out. There are also some properies that are commonly seen
throughout the geometric algebras we commonly deal with that are \textit{not} preserved across the
all geometric algebras.

\subsection{Properties shared across geometric algebras}

\subsubsection{Associativity of the geometric product}

All geometric algebras are associative, so the order in which pairs of 

\subsubsection{Anticommutativity of the wedge product}

\subsection{Properties not shared across geometric algebras}

\subsubsection{Commutativity of the geometric product}

Only a few geometric algebras are commutative - most notably, the Clifford algebra isomorphic to the
imaginary numbers, $\ClR{0,1}$.

\subsubsection{$k$-vectors are not always blades}

For Clifford algebras up to 3 dimensions, any $k$-vector may be written as the product of basis
blades. However, in high dimensions, this is not necessarily true. The most commonly cited example
is the $k$-vector $e_1 e_2 + e_3 e_4$. 

\subsubsection{There may be zero divisors}

\textit{Zero divisors} are nonzero elements of an algebra which can result in a zero product when
multiplied by a nonzero element in the algebra. In the algebra of physical space, the product
$(1 + e_1)(1 - e_1)$ equals zero.
